Efficient Computation of Extended Sources

### 2021-05-22 repro/papers/sseprecond/paper.tex
interim

\plot{daltplh0tr81}{width=\textwidth}{Overplot of traces 81 ($x=3600$)
  from gathers shown in \ref{fig:drecplh0} (blue), \ref{fig:daltplh0}
  (red).}
\section{Approximate Inversion}

@@

Computation of $\Lambda^{\pm}_{z_s}$ will prove a critical step in an
efficient iterative solution of the linear subproblem
\ref{eqn:esis}. Direct computation, for instance by solving
\ref{eqn:sawe2} and reading off $P_sv^{\pm}_{2,z}$, turns out to be
numerically ill-behaved. An accurate computation is available via the
computation of a {\em parametrix} (high frequency approximate inverse)
for $S^{\pm}_{z_s,z_r}$. This parametrix also justifies the use of the
symmetric part of $\Lambda^{\pm}_{z_s}$ as a preconditioner. In this
section I will explain the parametrix construction, and also show how
to extract the symmetric part of $\Lambda^{\pm}_{z_s}$.

The key ingredient in this {\em time reversal} construction is a
{\em local energy decay} assumption. Define the local energy in the
acoustic field $(p,\bv)$ at
time $t$ in a region $\Omega \subset \bR^3$ by imiting the integration
in the definition of total energy \ref{eqn:defae0} to $\Omega$:
\begin{equation}
  \label{eqn:defaeloc}
  E_{\Omega}[(p,\bv)](t) = \frac{1}{2} \int_{\Omega} \,d\bx \, \left(\frac{p(\bx,t)^2}{\kappa(\bx)} + \rho(\bx) |\bv(\bx,t)|^2\right)
\end{equation}

We assume that $(p^+,\bv^+)$ is {\em downgoing} in the half-space
$\Omega = \{(x,y,z): z \le z_r\}$: that is, there exists a solution of
the acoustic system $(\tilde{p}^+,\tilde{\bv}^+)$ in space-time $\bR^4$
for which
\begin{itemize}
\item[1. ] $p^+ \approx \tilde{p}^+, \bv^+ \approx \tilde{\bv}^+$ in
  the slab $\Gamma = \{(x,y,z): z_s \le z \le z_r\}$, and
\item[2. ] The energy in $(\tilde{p}^+,\tilde{\bv}^+)$ decays in
  forward time, in the half-space $\Omega$:  $E_{\Omega}[(\tilde{p}^+,\tilde{\bv}^+)](t) \rightarrow 0$
  as $t \rightarrow \infty$.
\end{itemize}

Conditions for local energy decay have been widely studied in various
settings \cite[]{MorawetzPhillipsLax:05,Hristova:09}. For the half-space
geometries considered here, local energy does not decay in general,
because of the existence of ``trapped'' rays traveling parallel to the
boundary. Instead, high frequency data associated with rays crossing
the boundary $\{z=z_r\}$ transversally, and never returning to the slab, have
the local energy decay property.

The downgoing constraint has mostly been studied in relation to
seismology, mostly as a ray-theoretic concept: downgoing fields are
concentrated in phase space along rays for which depth $z$ is
increasing with $t$. The version above is stated without reference to
ray theory, but the setting of this paper allows its use and in a
later section I will explicitly construct the downgoing approximation
under appropriate ray-theoretic conditions. The numerical examples will
illustrate the necessary assumptions. When ray-theory is appropriate,
the downgoing constraint can be {\em ex post facto} enforced via dip
filtering. For example, the application of the ideas developed here to
diving wave inversion will require such a filtering step.

An immediate consequence of the energy decay assumption is that in the
slab $\Gamma$, the field $(p^+,\bv^+)$ approximates the solution of an
anti-causal evolution equation. For any choice of $T$, define
$(\tilde{p}^-,\tilde{\bv}^-)$ to be the solution in the half-space
$\Omega$ of
\begin{eqnarray}
\label{eqn:revawe}
  \frac{1}{\kappa}\frac{\partial \tilde{p}^-}{\partial t} & = & - \nabla \cdot \tilde{\bv}^-, \nonumber \\
  \rho\frac{\partial \tilde{\bv}^-}{\partial t} & = & - \nabla \tilde{p}^-,\nonumber \\
  \tilde{p}^- & =& 0,  \mbox{ for } t=T,\nonumber\\ 
  \tilde{\bv}^- & = & 0 \mbox{ for } t=T,\nonumber\\
  P_r\tilde{p}^- &=& P_rp^+ \mbox { for } t \le T . 
\end{eqnarray}
As will be shown later, under the downgoing assumtion, $P_r p^+
\approx P_r \tilde{p}^+$.
Therefore $(\delta p,\delta \bv)=(\tilde{p}^+-\tilde{p}^-,\tilde{\bv}^+-\tilde{\bv}^-)$
differs negligibly from the solution of
\begin{eqnarray}
\label{eqn:deltaawe}
  \frac{1}{\kappa}\frac{\partial \delta p}{\partial t} & = & -
                                                               \nabla
                                                               \cdot
                                                               \delta \bv, \nonumber \\
  \rho\frac{\partial \delta \bv}{\partial t} & = & - \nabla \delta p,\nonumber \\
  \delta p & =& \tilde{p}^+,  \mbox{ for } t=T,\nonumber\\ 
  \delta \bv & = & \tilde{\bv}^+ \mbox{ for } t=T,\nonumber\\
  P_r\delta p & = & 0,
\end{eqnarray}
in $\Omega \times \bR$. Since
$E_{\Omega}[(\tilde{p}^+,\tilde{\bv}^+)](T) \rightarrow 0$ as $T
\rightarrow \infty$, the usual energy estimate implies that the energy $\delta
E$ in this field,
\[
  \delta E_{\Omega}(t) = \frac{1}{2} \int_{\Omega} \,d\bx \,
 \left(\frac{(\delta p)^2}{\kappa} + \rho |\delta \bv|^2\right)
\]
satisfies $\delta E_{\Omega}(t) \rightarrow 0$ as $T \rightarrow
\infty$. This decay does not generally imply that $P_s\delta p, P_s \delta \bv)
\rightarrow 0$ (for instance) as $T \rightarrow \infty$, since the trace operator
$P_s$ is not continuous, but it does so imply for purely downgoing
fields, as will be demonstrated later, so
\begin{equation}
  \label{eqn:approxinv0}
  \lim_{T \rightarrow \infty} P_s\tilde{v}_z^- \approx \lim_{z \rightarrow
    0^+}\tilde{v}_z^+ \approx \lim_{z \rightarrow 0^+}v^+_z
\end{equation}

>>>>

\plot{daltplh0tr81}{width=\textwidth}{Overplot of traces 81 ($x=3600$)
  from gathers shown in \ref{fig:drecplh0} (blue), \ref{fig:daltplh0}
  (red).}

\section{Approximate Inversion}

Computation of $\Lambda^{\pm}_{z_s}$ will prove a critical step in an
efficient iterative solution of the linear subproblem
\ref{eqn:esis}. Direct computation, for instance by solving
\ref{eqn:sawe2} and reading off $P_sv^{\pm}_{2,z}$, turns out to be
numerically ill-behaved. The relation \ref{eqn:snull} provides and
alternative approach, {\em provided} that an accurate approximate
inverse to $S^+_{z_s,z_r}$ is available. For instance, the first row
of the vector equation \ref{eqn:snull} reads (slightly rearranged):
\[
\Pi_0{\cal S}^+_{z_s,z_r}\Pi_1 f_s \approx \Pi_0{\cal
  S}^+_{z_s,z_r}Pi_0\Lambda^+_{z_s}f_s
\]
\begin{equation}
  \label{eqn:lamidea}
  = S^++_{z_s,z_r}\Lambda^+_{z_s}f_s.
\end{equation}
An approximate inverse for $S^++_{z_s,z_r}$ would permit solution of
this equation for $\Lambda^+_{z_s}f_s$.

In this section I will explain how to build the approximate inverse.
The key ingredient in this {\em time reversal} construction is a
{\em local energy decay} assumption. Define the local energy in the
acoustic field $(p,\bv)$ at
time $t$ in a region $\Omega \subset \bR^3$ by limiting the integration
in the definition of total energy \ref{eqn:defae0} to $\Omega$:
\begin{equation}
  \label{eqn:defaeloc}
  E_{\Omega}[(p,\bv)](t) = \frac{1}{2} \int_{\Omega} \,d\bx \, \left(\frac{p(\bx,t)^2}{\kappa(\bx)} + \rho(\bx) |\bv(\bx,t)|^2\right)
\end{equation}
The necessary assumption is: local energy decay in the half-space
$\Omega = \{(x,y,z): z \le z_r\}$. The local energy in $\Omega$ of the
causal solution $(p^+,\bv^+)$ of system \ref{eqn:awepm} cannot decay,
since the source on $z=z_s<z_r$ radiates in both positive and negative
$z$ in forward time. That obstacle is resolved by a redefinition of
the ``downgoing'' concept. I will call  downgoing if there is a
source-free acoustic field $(\tilde{p}^+,\tilde{\bv}^+)$ in all of space-time for which
\begin{itemize}
\item[1. ] $p^+ \approx \tilde{p}^+, \bv^+ \approx \tilde{\bv}^+$ in
  the slab $\{(x,y,z,t): z_s \le z \le z_r\}$, and
\item[2. ] The energy in $(\tilde{p}^+,\tilde{\bv}^+)$ decays in
  forward time, in the half-space $\Omega$:  $E_{\Omega}[(\tilde{p}^+,\tilde{\bv}^+)](t) \rightarrow 0$
  as $t \rightarrow \infty$.
\end{itemize}
Conditions for local energy decay have been widely studied in various
settings, for instance obstacle scattering
\cite[]{MorawetzPhillipsLax:05}. \cite{Pauen2000NonTr-6079 gives a
comprehensive overview of this subject. For smooth (non-scattering)
acoustic parameters, as is assumed in this paper, the crux of the
matter is that rays carrying high-frequence energy exit the half-space
$\Omega$ in finite time. For the half-space
geometries considered here, local energy does not decay in general,
because of the existence of ``trapped'' rays traveling parallel to the
boundary and remaining interior to $\Omega$ for all time. If an
acoustic field is downgoing in the usual sense, that is, its
high-frequency energy is concentrated on rays that exit $\Omega$
within some common maxiumum time, then the results cited above show
that local energy in $\Omega$ decays (exponentially for $n=3$).

Starting from this observation, it is possible to show that if the
source vector $(h_s,f_s)^T$ has the property termed ``downgoing'' in
preceding sections, that is, emanates high-frequency energy only along
rays that leave $\Omega$ within a common maximum time, then the
solution $(p^+,\bv^+)$ of \ref{eqn:awepm} is downgoing in the sense
defined here. All of our numerical examples behave as one would expect
for downgoing fields.

An immediate consequence of the downgoing assumption stated above is that in the
slab $z_s<z<z_r$, the field $(p^+,\bv^+)$ approximates the solution of an
anti-causal evolution equation. For any choice of $T$, define
$(\tilde{p}^-,\tilde{\bv}^-)$ to be the solution in the half-space
$\Omega$ of
\begin{eqnarray}
\label{eqn:revawe}
  \frac{1}{\kappa}\frac{\partial \tilde{p}^-}{\partial t} & = & - \nabla \cdot \tilde{\bv}^-, \nonumber \\
  \rho\frac{\partial \tilde{\bv}^-}{\partial t} & = & - \nabla \tilde{p}^-,\nonumber \\
  \tilde{p}^- & =& 0,  \mbox{ for } t=T,\nonumber\\ 
  \tilde{\bv}^- & = & 0 \mbox{ for } t=T,\nonumber\\
  P_r\tilde{p}^- &=& P_rp^+ \mbox { for } t \le T . 
\end{eqnarray}
As will be shown later, under the downgoing assumption, $P_r p^+
\approx P_r \tilde{p}^+$.
Therefore $(\delta p,\delta \bv)=(\tilde{p}^+-\tilde{p}^-,\tilde{\bv}^+-\tilde{\bv}^-)$
differs negligibly from the solution of
\begin{eqnarray}
\label{eqn:deltaawe}
  \frac{1}{\kappa}\frac{\partial \delta p}{\partial t} & = & -
                                                               \nabla
                                                               \cdot
                                                               \delta \bv, \nonumber \\
  \rho\frac{\partial \delta \bv}{\partial t} & = & - \nabla \delta p,\nonumber \\
  \delta p & =& \tilde{p}^+,  \mbox{ for } t=T,\nonumber\\ 
  \delta \bv & = & \tilde{\bv}^+ \mbox{ for } t=T,\nonumber\\
  P_r\delta p & = & 0,
\end{eqnarray}
in $\Omega \times \bR$. Since
$E_{\Omega}[(\tilde{p}^+,\tilde{\bv}^+)](T) \rightarrow 0$ as $T
\rightarrow \infty$, the usual energy estimate implies that the energy $\delta
E$ in this field,
\[
  \delta E_{\Omega}(t) = \frac{1}{2} \int_{\Omega} \,d\bx \,
 \left(\frac{(\delta p)^2}{\kappa} + \rho |\delta \bv|^2\right)
\]
satisfies $\delta E_{\Omega}(t) \rightarrow 0$ as $T \rightarrow
\infty$. This decay does not generally imply that $P_s\delta p, P_s \delta \bv)
\rightarrow 0$ (for instance) as $T \rightarrow \infty$, since the trace operator
$P_s$ is not continuous, but it does so imply for purely downgoing
fields, as will be demonstrated later, so
\begin{equation}
  \label{eqn:approxinv0}
  \lim_{T \rightarrow \infty} P_s\tilde{v}_z^- \approx \lim_{z \rightarrow
    0^+}\tilde{v}_z^+ \approx \lim_{z \rightarrow 0^+}v^+_z
\end{equation}

@@
